How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-10-02 analysis/paper/manuscript.Rmd

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Fleur Hierink:
      institute: [Geneva]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Jean-Michel Heraud:
      institute: [IPMviro]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: World Health Organization, Madagascar
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Geneva: University of Geneva
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
nocite: '@*'
output:
 bookdown::word_document2:
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: no
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
bibliography_main: ["bib/refs_all.bib"]
bibliography_supp: ["bib/refs_pkgs.bib"]
---

```{r setupmain, include=FALSE}
# Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(sf)
library(glue)
library(here)
library(magrittr)
opts_chunk$set(echo = FALSE, warning = FALSE, out.width="80%")

@@

## data
ctar_metadata <- fread(here("data-raw/out/clinics/ctar_metadata.csv"))
district_bites <- fread(here("analysis/out/bites/district_bites.csv"))
mora_bites <- fread(here("analysis/out/bites/mora_bites.csv"))
mada_communes <- st_read(here("analysis/out/shapefiles/mada_communes.shp"))
mada_districts <- st_read(here("analysis/out/shapefiles/mada_districts.shp"))

## All preds (for abstract stats etc)
natl_preds <- fread(here("analysis/out/preds/natl_preds.gz"))
natl_base <- natl_preds[scenario == 0]

@@

## candidate points
csbs <- fread(here("data-raw/out/clinics/csb2.csv"))
clin_per_comm <- fread(here("data-raw/out/clinics/clinic_per_comm.csv"))
clin_per_dist <- fread(here("data-raw/out/clinics/clinic_per_dist.csv"))
missing_comms <- fread(here("analysis/out/stats/missing_comms.csv"))
comms_no_clins <- nrow(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)

max_added <- max(natl_preds$scenario)
natl_preds %>%
  group_by(scale, data_source) %>%
  arrange(scenario) %>%
  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
trends <- read.csv(here("analysis/out/stats/trends_adminpreds.csv"))

@@

# ABSTRACT

## Background
For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. A target to eliminate deaths due to canine rabies by the year 2030 has been set forth by global health partners, and in 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly expand PEP availability and reduce rabies burden. Given the potential for countries to provision PEP more widely, we estimate the impact of access to PEP on burden and explore how expanding access can reduce burden in Madagascar, where current access to PEP is highly limited (to just 31 clinics across the country).

## Methodology & Principal Findings
We use travel times to the closest ARMC as a proxy for access to PEP. We find that travel times are a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes burden sub-nationally. We estimate an annual burden of `r round(natl_base$deaths_mean[1], -1)` (95% PI: `r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with current PEP provisioning averting an additional `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) deaths. Under these assumptions, we estimate that expanding PEP provisioning to one clinic per district could reduce the burden of rabies deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`. However, our results suggest that even with all major health centers in the country provisioning PEP (N = `r max(natl_preds$scenario)` additional clinics), we could still see substantial mortality due to rabies. Our estimates are most sensitive to assumptions of underlying rabies exposure incidence, but the qualitative patterns of the impacts of travel times and expanded access on burden are robust to these assumptions.

## Conclusions & Significance
PEP is effective at preventing deaths and in the absence of rabies data, targeting expansion to populations currently underserved may be the most equitable way to provision PEP. Data on rabies incidence will be neccessary to refine these estimates and to better target interventions. While expanding access could save many more lives, our results suggest that improved outreach and awareness is also needed in addition to improved vaccine provisioning and surveillance systems. A GAVI investment can propel countries forward on this path combining these life saving interventions.

# AUTHOR SUMMARY 
Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low and middle income countries where people have limited access to both human (post-exposure prophylaxis or PEP) and dog rabies vaccine. Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by the year 2030, and recently, GAVI, a organization that provides funding for immunization programs to eligible countries, added human rabies vaccine to it's funding agenda. In this study, we model estimates of reported bite incidence as a function of travel times and estimates of rabies exposure incidence to extrapolate the burden of human rabies given the current availability of PEP across the country. We find that PEP currently averts approximately `r round(natl_base$averted_mean[1], -2)` deaths, but that burden remains high, particularly in areas with high travel times to a clinic. Expanding PEP availability to more clinics could significantly reduce the current burden of rabies deaths, but our results suggest that expansion alone will not eliminate deaths due to rabies under current levels of reporting. Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other countries like it, forwards on the path to elimination.

@@

# INTRODUCTION

@@

Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans. While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018]. However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].

In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero]. Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS]. From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.

A primary challenge in expanding access effectively is the lack of data on rabies exposures and deathss in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in the most vulnerable and remote communities [@taylor2017difficulties]. Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites and a decision tree framework to extrapolate deaths, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country [@rajeevHealthcareUtilizationProvisioning2018]. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control of rabies in dog populations and the disease appears to be endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011]. Due to the spatially restricted nature of PEP distribution points and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country. Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2018]. Here, we revise estimates of burden by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.

# METHODS

## Estimating travel time to closest ARMC {#methods-ttimes}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B.

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP at an approximately 1 x 1 km scale. We then extracted the mean travel times as well as the mean of travel times weighted by population for the district and commune (the administrative unit below the district) level. We also compared these metrics to euclidean distance, taking the minimum distance from the administrative unit centroid to any ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

We compared these travel time and distance estimates to driving time data collected by IPM during field missions and patient reported travel times from a subset of patients reporting to the Moramanga ARMC (see Fig S1.2 for raw data).

## Estimating average annual bite incidence {#methods-bitedata}
We used two datasets on bite patients reporting to ARMC:

1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually, which included data on the date of reporting and were resolved to the district level (i.e. the district where the patient resided).
2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level. We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period. However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1)). We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic. Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.

We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period. Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day. We did not have the date or month when vials were provisioned, so we compared estimated demand to the total vials provisioned across this four year period.

At both the commune and district scale, we assigned clinic catchments by determining which clinic was closest in terms of travel times for the majority of the population within the administrative unit. For the national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level. In order to more directly compare estimates between these two datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$) so that:

$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0_j)}P_j$$

As travel times are correlated with human population size (Figure S3.1), we also compared how well bites were predicted by human population size alone, and in combination with travel times. For the models with population size, we used a model framework with either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors of bites for the Moramanga data, the national data with covariates at the district level, and the national data with covariates at the commune level.

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma_0)$) where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept). The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.

We tested whether the catchment random effect was capturing overdispersion in the data (i.e. variance greater than the mean--the expectation of variance given a poisson distribution) rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:

$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j + \sigma_e)}P_i$$
where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters and for a given model.

>>>>

---
title: "How geographic access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Hélène Guis:
      institute: [IPMepi, cirad, astre, fofifa]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Fleur Hierink:
      institute: [Geneva]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - José Nely:
      institute: [smeetn]
  - Jean-Michel Heraud:
      institute: [IPMviro]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
      equal_contributor: "yes"
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
      equal_contributor: "yes"
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
      equal_contributor: "yes"
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - cirad: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - astre: ASTRE, Univ Montpellier, CIRAD, INRAE, Montpellier, France 
  - fofifa: FOFIFA-DRZVP, Antananarivo, Madagascar6 World Health Organization, Madagascar
  - smeetn: Service contre les Maladies Endémo-épidémiques et Tropicales  Négligées, Ministère de la Santé Publique, Antananarivo, Madagascar
  - OMS: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Geneva: Institute of Global Health, Faculty of Medicine, University of Geneva, Geneva, Switzerland
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur du Cambodge (previously Epidemiology and Clinical Research Unit at Institut pasteur de Madagascar)
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
nocite: '@*'
link-citations: no
output:
  bookdown::word_document2:
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/reference_supplement.docx
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: no
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
#bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib", "bib/refs_supextra.bib"]
bibliography_main: ["bib/refs_all.bib"]
bibliography_supp: ["bib/refs_pkgs.bib"]
bibliography_supextra: ["bib/refs_supextra.bib"]
---

```{r setupmain, include=FALSE}
# Libraries
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(sf)
library(glue)
library(here)
library(magrittr)
source(here("R", "utils.R"))

opts_chunk$set(echo = FALSE, warning = FALSE, 
                message=FALSE, out.width="80%")

@@

# data
ctar_metadata <- fread(here("data-raw/out/clinics/ctar_metadata.csv"))
district_bites <- fread(here("analysis/out/bites/district_bites.csv"))
mora_bites <- fread(here("analysis/out/bites/mora_bites.csv"))
mada_communes <- st_read(here("analysis/out/shapefiles/mada_communes.shp"))
mada_districts <- st_read(here("analysis/out/shapefiles/mada_districts.shp"))

# All preds (for abstract stats etc)
natl_preds <- fread(here("analysis/out/preds/natl_preds.gz"))
natl_base <- natl_preds[scenario == 0]

@@

# candidate points
csbs <- fread(here("data-raw/out/clinics/csb2.csv"))
clin_per_comm <- fread(here("data-raw/out/clinics/clinic_per_comm.csv"))
clin_per_dist <- fread(here("data-raw/out/clinics/clinic_per_dist.csv"))

comms_no_csb2 <- nrow(clin_per_comm[type == "CSB1"])

comms_no_clins <- 
  nrow(mada_communes) - 
  length(unique(c(clin_per_comm$commcode, 
                  clin_per_dist$commcode, 
                  ctar_metadata$commcode)))

csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)
csb2_max <- sum(csbs_max$type == "CSB2")
csb1_max <- sum(csbs_max$type == "CSB1")
max_added <- max(natl_preds$scenario)
natl_preds %>%
  group_by(scale, data_source) %>%
  arrange(scenario) %>%
  mutate_at(vars(-c(group_cols(), scenario)), function(x) x/x[1]) -> props
trends <- read.csv(here("analysis/out/stats/trends_adminpreds.csv"))

@@

# Abstract

## Background
Post-exposure prophylaxis (PEP) is highly effective at preventing human rabies deaths, however access to PEP is limited in many rabies endemic countries. The 2018 decision by Gavi to add human rabies vaccine to its investment portfolio should expand PEP availability and reduce rabies deaths. We explore how geographic access to PEP impacts the rabies burden in Madagascar and the potential benefits of improved provisioning.

## Methodology & Principal Findings
We use travel times to the closest clinic providing PEP (N=31) as a proxy for access. We find that travel times strongly predict reported bite incidence across the country. Using resulting estimates in an adapted decision tree framework we extrapolate rabies deaths and reporting and find that geographic access to PEP shapes burden sub-nationally. We estimate 960 human rabies deaths annually (95% Prediction Intervals (PI):`r round(natl_base$deaths_lower[1], -1)` - `r round(natl_base$deaths_upper[1], -1)`), with PEP averting an additional 800 deaths (95% PI: `r round(natl_base$averted_mean[1], -1)` (95% PI: `r round(natl_base$averted_lower[1], -1)` - `r round(natl_base$averted_upper[1], -1)`) each year. Under these assumptions, we find that expanding PEP to one clinic per district could reduce deaths by `r glue("{trans_perc((1 - props$deaths_mean[props$scenario == 83 & props$scale == 'Commune']))}%")`, but even with all major health centers provisioning PEP (`r max(natl_preds$scenario)` additional clinics), we still expect substantial rabies mortality. Our quantitative estimates are most sensitive to assumptions of underlying rabies exposure incidence, but qualitative patterns of the impacts of travel times and expanded PEP access are robust.

## Conclusions & Significance
PEP is effective at preventing rabies deaths, and in the absence of strong surveillance, targeting underserved populations may be the most equitable way to provision PEP. Our framework could be used to guide PEP expansion and improve targeting of interventions in similar endemic settings where PEP access is geographically restricted. While better PEP access should save many lives, improved outreach and surveillance is needed and if rolled out with Gavi investment could catalyse countries' progress towards achieving zero rabies deaths.

# Author Summary 
Canine rabies causes an estimated 60,000 deaths each year across the world, primarily in low- and middle-income countries where people have limited access to both human vaccines (post-exposure prophylaxis or PEP) and dog rabies vaccines. Given that we have the tools to prevent rabies deaths, a global target has been set to eliminate deaths due to canine rabies by 2030, and recently, Gavi, a multilateral organization that aims to improve access to vaccines in the poorest countries, added human rabies vaccine to it’s portfolio. In this study, we estimated reported bite incidence in relation to travel times to clinics provisioning PEP, and extrapolate human rabies deaths in Madagascar. We find that PEP currently averts around `r round(natl_base$averted_mean[1], -2)` deaths each year, but that the burden remains high (`r round(natl_base$deaths_mean[1], -2)` deaths/ year), particularly in remote, hard-to-reach areas. We show that expanding PEP availability to more clinics could significantly reduce rabies deaths in Madagascar, but our results suggest that expansion alone will not eliminate deaths. Combining PEP expansion with outreach, surveillance, and mass dog vaccination programs will be necessary to move Madagascar, and other Low- and Middle-Income countries, forward on the path to rabies elimination.

@@

# Introduction

@@

Canine rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans. While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid animal [@RabiesVaccinesWHO2018]. However, access to human vaccine is limited in many countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].

In 2015, a global framework to eliminate deaths due to canine rabies by 2030 ('Zero by 30') through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero]. Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS]. From 2021, Gavi-eligible countries should be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies.

A primary challenge in expanding access effectively is the lack of data on rabies exposures and deaths in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. Deaths due to rabies are often severely underreported, with many people dying outside of the health system, often in remote  and marginalized communities [@taylor2017difficulties]. Instead of directly measuring rabies deaths, the majority of rabies burden studies use bite patient data on reported bites at clinics provisioning PEP and a decision tree framework to extrapolate deaths, assuming that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) is proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting to clinics for PEP is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2019]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP to 30 Ministry of Health clinics, in addition to it's own vaccine clinic, where PEP is available at no direct cost to patients [@rajeevHealthcareUtilizationProvisioning2019]. Other than at these 31 anti-rabies medical centers (ARMC), PEP is not available at any other public clinics or through the private sector. In addition, there is limited control of rabies in dog populations and the disease is endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011]. Due to the spatially restricted nature of PEP provisioning and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country. Previously, we estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2019]. Here, we provide revised estimates of human rabies deaths by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths. This framework may usefully apply to other countries where PEP availability is currently geographically restricted in considering how to most effectively and equitably provision these life-saving vaccines.

# Methods

## Estimating geographic access to ARMC {#methods-ttimes}
To estimate mean and population weighted travel times to the nearest clinic at the scale of administrative units, we used two rasters: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale (Fig S1.1A) and 2) the population estimates from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, Fig S1.1B), which we aggregated to the friction surface.

From GPS locations of the 31 ARMCs we estimated the travel time to the nearest clinic at an approximately 1 x 1 km scale as described in [@weissGlobalMapTravel2018]. We then extracted the mean and population-weighted mean travel times for the district and commune (the administrative unit below the district), and euclidean distance, i.e. the minimum distance from the administrative unit centroid to any ARMC. We used shapefiles from the UN Office for the Coordination of Humanitarian Affairs for the administrative boundaries (as of October 31, 2018). We compared travel times and distance estimates to driving times collected by IPM during field missions and patient reported travel times from a subset of Moramanga ARMC patients (see Fig S1.2 for raw data).

## Estimating bite incidence {#methods-bitedata}
We used two datasets on bite patients reporting to ARMC:

1)	A national database of individual bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These forms were submitted with frequencies ranging from monthly to annually, included the patient reporting date and were resolved to the district level (patient residence).

2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga District ARMC resolved to commune level.

For the national data, some clinics did not submit any data, or had substantial periods (months to a whole year), with no submitted data. To correct for this, we exclude periods of 15 consecutive days with zero submitted records (see Supplementary Appendix, section S2). For each clinic we divided the total number of bites reported in a given year by the estimated proportion of forms which were not submitted (under-submission). Due to yearly variation in submissions, we took the average of annual bite incidence estimates aggregated to district level. We validated this approach by comparing estimated vial demand given the total reported bites corrected for under-submission to vials provisioned to clinics for 2014-2017 (see Supplementary Appendix, section S2).
At both the commune and district level, we assigned clinic catchments by determining which were closest in terms of travel times for the majority of the population within the administrative unit. For national data, we excluded any districts in a catchment of a clinic which submitted less than 10 forms and any years for which we estimated less than 25% of forms were submitted.

## Modeling reported bite incidence {#methods-bitemods}

We modeled the number of reported bites as a function of travel time ($T$) using a Poisson regression:

$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$

where $y_i$ is the average number of bites reported to a clinic annually and $\mu_i$ the expected number of bite patients presenting at the ARMC as a function of travel time ($T_i$) and human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune). We fit this model to both the national data (district level) and the Moramanga data (commune level). To more directly compare estimates between datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$):

$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ \ + \ \beta_{0j})}P_j$$

As travel times are correlated with population size (Fig S3.1), we also compared how well bites were predicted by population size alone, and in combination with travel times. For the models with population size, we removed the offset and used either population size alone ($\mu_i = e^{(\beta_pP_i + \beta_0 )}$) or population size and travel times ($\mu_i = e^{(\beta_{t}T_i + \beta_pP_i + \beta_0 ))}$) as predictors.

For the models fit to the national data, we also modeled variation between clinics with a catchment random effect: $B_{0,k} \sim norm(\mu,\ \sigma_0)$), where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept.

We tested whether the catchment random effect captured overdispersion in the data (i.e. variance > mean – the expectation given a Poisson distribution) rather than any catchment specific effects by extending these models with an overdispersion parameter, $\sigma_e$ [@gelman2006data]:

$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j \ + \ \sigma_e)}P_i$$

where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters for a given model. 
We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’ [@R-rjags]. We used model estimates to generate fitted and out-of-fit predictions, and examined the sensitivity of estimates to adjustments for under-submission of forms (Supplementary Appendix, section S3).

## Modeling human rabies deaths {#methods-burden}
We estimate rabies deaths as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1). To model uncertainty in parameter estimates we used triangular distributions, as with previous studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual exposure incidence per administrative unit, and  $p_{rabid}$, the proportion of reported bites that are rabies exposures.

```{r fig 1, out.width="100%"}
include_graphics(here("data-raw/misc/figM1/figM1_final.tiff"))
```
